---
title: "Residential Appeal SOPs"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align='center', warning = FALSE, fig.align = "center")

# Load the necessary libraries
library(ggpubr); library(ccao); library(DBI)
library(ggplot2); library(gridExtra); library(htmltools)
library(htmlwidgets); library(jsonlite); library(knitr)
library(leaflet); library(lubridate); library(odbc)
library(readr); library(readxl); library(RSocrata)
library(scales); library(sf); library(stargazer)
library(testit); library(tidyr); library(tidyverse)
library(writexl); library(ccao); library(DT)
library(margins)

# Load
load('res_appeals_graph_data_1.Rda')
load('res_appeals_graph_data_2.Rda')

# need in-line citation data

```

# Opportunity
The Cook County Assessor’s Office (CCAO) is committed to fair, transparent, ethical assessments. The Data Science Department (DS) proposes these Standard Operating Procedures (SOPs) to govern the CCAO's residential property appeals process starting 2021. These SOPs provide comprehensive, clear, consistent guidelines to residential appeals analysts and other CCAO staff processing residential appeals. These SOPs will also be available to the public so that taxpayers can understand and anticipate appeal decisions before filing their appeal. These SOPs will help de-mystify the residential appeals process, reducing potential for unethical behavior and increasing transparency and fairness.

# Motivation

This section discusses the motivation for adopting these SOPs. Transparency [requires us](https://gitlab.com/ccao-data-science---modeling/documentation/wiki_content/-/blob/master/handbook/mission%20vision%20values.md#transparency) to release both the policy itself, as well as the motivation and intended outcomes of the policy. To the degree possible, we have written these SOPs so that they are understandable to the typical taxpayer.

## Fairness

Illinois property taxes operate according to a levy-based system. In this system, local governments establish the amount of money they want to raise from property taxes, and the assessment process determines the distribution of that tax burden. During a re-assessment, assessors mail initial values to taxpayers, who then have the opportunity to appeal. Successful appeals _do not deprive the government of tax revenue._ Instead, successful appeals shift the burden of a pre-determined levy from taxpayers with successful appeals to those without successful appeals. Under this system, the appeals process has inherent consequences for fairness.

In most Illinois counties, between 5/% and 10/% of residential property owners appeal their assessments **FACT CHECK PLEASE.** In Cook County, 35/% of residential taxpayers appealed their 2019 triennial re-assessment. As a consequence, residential appeals account for a large portion of the agency's workload; in 2019, the CCAO processed more than 100,000 residential appeals of triennial re-assessments. Residential assessments significantly impact of the distribution of property taxes in Cook County; during the past three re-assessments, residential appeals reduced the size of jurisdiction's total assessed values by 3/% to 6/%. The fairness of the appeals process is a particularly important consideration in Cook County because of the large volume of appeals.

### Appeals make the system less fair overall

The CCAO did not perform sales ratio studies prior to the Kaegi administration. In response to recommendations [(pg. 17)](https://prodassets.cookcountyassessor.com/s3fs-public/reports/PCS_Cook-County_Report_final.pdf) from the International Association of Assessing Officer (IAAO) DS developed a robust set of diagnostic tools to determine whether residential assessments were fair and accurate. The residential valuation reports [released with each township mailing](https://www.cookcountyassessor.com/valuation-reports), and the [residential diagnostic report](https://gitlab.com/ccao-data-science---modeling/reports/assessment-diagnostic-report) consistently raise questions about the accuracy and fairness of the CCAO's residential appeals decisions.

A number of credible third party reports have also found that residential appeals reduce fairness in assessments. In 2016, researchers at the [University of Chicago](https://harris.uchicago.edu/files/inline-files/Ross%20-%20Vertical%20Equity%20in%20Cook%20County%20-%20CMF%20-%20Final.pdf), in collaboration with [ProPublic](https://www.propublica.org/series/the-tax-divide) and [The Chicago Tribune](https://www.chicagotribune.com/investigations/ct-tax-divide-investigation-20180425-storygallery.html) found that appeals decrease fairness in assessments overall. In 2018, a [Civic Consulting Alliance report](https://prodassets.cookcountyassessor.com/s3fs-public/reports/CivicConsultingAllianceStudyResults.pdf) found the same. There is ample reason to believe that the CCAO's residential appeals process does not reliably or consistently improve fairness in assessments overall. 

### Taxpayers with more valuable homes are more likely to appeal, be successful

The two primary drivers of unfairness in the appeals process are the self-selective nature of appeals, and the way in which our office processes appeals. Since appeals are initiated by taxpayers, taxpayers with more valuable homes have a greater incentive to appeal. Our office disproportionatly processes appeals from higher-value homes. Partly as a result of this self selection, higher value homes are also more likely to win their appeal, and win a larger reduction as a percentage of value on appeal.

``` {r, appeal_by_value}

g1 <- ggplot(subset(gd1, Period=='2019' & Statistic != 'Mean reduction %')
             , aes(x=`First Pass Market Value`, y=Rate))+
  geom_bar(stat='identity', aes(fill=Statistic), position = "dodge")+
  scale_y_continuous(labels = function(x) paste0(x*100, "%")) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90),
        legend.position = 'bottom') +
  scale_fill_manual(values=c("#29428D","#FFAF00")) +
  labs(title = "Appeal rate & success rate",
  subtitle = '2019 re-assessment')+
  ylab(' ')
g2 <- ggplot(subset(gd1, Period=='2019' & Statistic == 'Mean reduction %')
             , aes(x=`First Pass Market Value`, y=Rate))+
  geom_bar(stat='identity', aes(fill=Statistic), position = "dodge")+
  scale_y_continuous(labels = function(x) paste0(x*100, "%")) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90),
        legend.position = 'bottom') +
  scale_fill_manual(values=c("#81CA9C")) +
  labs(title = "Mean reduction for successful appeals", subtitle = '2019 re-assessment')+
  ylab(' ')+
  coord_cartesian( ylim = c(.04, .085))

ggarrange(g1, g2, ncol=2, nrow=1, common.legend = FALSE)

```

### Appeal decisions should be correlated with market signals

In the year following a reassessment, many properties transact. Each transaction can be used to test whether the initial assessment or the certified (post-appeal) assessment was closer to the sale price. The graph below shows three sets of data. The x-axis of the graph is the percent difference between a sale price in the year following a reassessment, and the initial value mailed to taxpayers that appealed. The y-axis shows the chances of having an appeal result in a reduction of value. Properties to the left of the black line are over-assessed, since their future sale price is less than their initial mailed value. Properties to the right of the black line are under-assessed, since their sale prices are greater than their mailed values. 

The green line simulates "accurate" appeal decisions. In this scenario, we gave analysts a 70% chance of identifying over-assessed properties as such. As a result, over-assessed properties' appeals are much more likely to be successful compared to those of under-assessed properties. The red line simulates the opposite scenario; appeals analysts simply cannot distinguish between over- and under-assessed properties. As a result, all appeals have an equal chance of winning, regardless of whether they are over- or under-assesed initially.

The orange line shows the average success rates for appeals of triennial assessments from 2015 to 2019. While the CCAO's appeal decisions are not entirely divorced from the market, there is significant room for improvement. For example, consider two properties, one over-assessed by 50%, and one under assessed by 50%. On average, the relative chances of a successful appeal for these properties are about 60% and 30% respectivly. Given such a large difference between the market and initial values of these properties, a larger difference in success chances would be preferred. We hope that these SOPs will increase the correlation between appeal decisions and market indicators. 


``` {r, appeal_accuracy}
graph_data_2 <- graph_data_2 %>% mutate(`Initial value difference`=`Initial value difference`/100)

g1 <- graph_data_2 %>%
  ggplot(aes(x=`Initial value difference`, y = `Appeal success rate`, group = `Decision type`, color =`Decision type`)) +
  geom_point() + geom_line() +
  geom_vline(xintercept = 0, linetype = 'dotted', size = 1) + 
  theme_minimal() +
  theme(legend.position = 'bottom') +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_color_manual(values = c('orange', 'green', 'red')) +
  labs(
    title = 'Appeal success rates'
    , subtitle = 'Simulated vs. real 2015 - 2019')

g2 <- graph_data_2 %>%
  dplyr::filter(`Decision type` != 'Simulated accurate decisions' & `Median reduction` > -.3) %>%
  ggplot(aes(x=`Initial value difference`, y = `Median reduction`
             , group = `Decision type`, color =`Decision type`)) +
  geom_point() + geom_line() +
  geom_vline(xintercept = 0, linetype = 'dotted', size = 1) + 
  theme_minimal() +
  theme(legend.position = 'bottom') +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(labels = scales::percent_format()) +
  geom_abline(intercept = 0, slope = 1, linetype = 'solid', size = 1, color = 'green') +
  scale_color_manual(values = c('orange', 'red')) +
  labs(
    title = 'Appeal reduction amounts'
    , subtitle = 'Simulated vs. real 2015 - 2019')

ggarrange(g1, g2, ncol=2, nrow=1, common.legend = TRUE)

```

## Transparency




When comparing appeal results for single-family appeals in 2019 and 2016, the data shows inconsistencies in appeal success rates. The linkage between the reason an appellant files a residential appeal, and the decision path an analyst uses to assign an appeal reason code are critical to our strategic values. Unfortunately, there is limited information available detailing what constitutes a successful appeal. 
For further details see [Comparing Residential Triennial Appeals in 2016 & 2019](https://gitlab.com/ccao-data-science---modeling/documentation/wiki_content/uploads/cf89ef12391221afbe4b1b6b8923c4af/Impact-of-Appeals-on-Vertical-Equity-in-a-Budget-Based-Property-Tax-Regime.html).


## Ethics

# Background

This section provides background information on the process of developing these SOPs.

## How do other agencies process residential appeals?

In 2020, the Data Science Department conducted a qualitative analysis on the transparency of rules to appeal. The public-facing websites of 10 counties in the United States was surveyed for well-defined rules on the residential appeals process and the criteria to win an appeal. When applying industry best practices themes, there were three consistent findings. First, none of the counties have a published single source of truth for the appeals process and win criteria. Second, six out of ten counties provided criteria on how to win an appeal, however there are inconsistencies due to the number of webpages and sources. Third, all counties provided varying degrees of information on reasons and rules to appeal, however the readability of this information does not perform well according to the basic standards of words per sentence, jargon, and length of text. Based on these observations, there is no industry standard for transparency of the appeals process. 
For further details see [Residential Appeals County Survey Data].

## What are the technical limitations on these SOPs?

### SmartFile


### Central data storage


### Other


## What are the technical assets for these SOPs

PINVAL

# Objectives and Key Results

SOPs create a rules-based approach to increase accuracy and predictability for analysts and appellants. We expect this approach to accomplish the following Key Results.

- Appeal decisions are consistent with market indicators.
- Appeal decisions are independent of one another, such that the order in which the decisions on two properties are made does not impact the decisions that are made.
- Analysts apply appeal reason codes that clearly indicate to everyone the reason(s) an appeal was granted or denied. Reason codes clearly correlate with these SOPs.
- Appellants apply these SOPs to their particular parcel in order to anticipate the outcome of their appeal.
- Citizens contribute to the development of these SOPs, which evolve over time to accomodate many edge cases.
- The cost of processing residential appeals remains constant, or declines; the SOPs do not impose an additional administrative burden on the agency.

## Key Performance Indicators

In order to evaluate whether these SOPs are accomplishing the Key Results, the CCAO will track a series of Key Performance Indicators (KPIs). 

- **Appeal SOPs are applied faithfully:** DS will randomly sample appeal results and investigate whether the SOPs were correctly applied. This is necessary to test the functional applicability of the SOPs and their implementation.
- **Appeal reason codes are clear and accurate:** DS will produce a report summarising appeal reason codes, success rates, and reduction amounts. This will help characterise the appeals process and determine whether codes need to be revised.
- **Board Appeals do not increase as a result of CCAO appeals:** Many taxpayer appeal both at the CCAO and at the Board of Review (BOR). DS will sample properties that had unsuccessful appeals at the CCAO, and successful appeals at the BOR. DS will identify all comparable properties for this set, and determine whether changes to each property's comparable set due to CCAO appeals created conditions for a successful BOR appeal.
- **Probability of success, amount of reduction: ** The success rate and mean reduction amount should be minimally correlated with property values. 
- **Appeal decision consistent with sales in following year:** Many properties that appeal transact after an appeal decision has been reached. This transaction provides a strong indicator for whether or not the appeal decision was consistent with the market. The CCAO will track both the probability of a successful appeal, and the mean reduction on appeal, across properties with sales in the following year. If appeal decisions are consistent with the market, there should be a strong correlation between the decision to grant an appeal and the amount of rediction, on one side, and the relative value of the sale price to the initial value on the other side.
- **The number of appeals filed decreases, appeals are more successful:** The rate of appeal should approach 10/%, while the rate of success should approach 90/%.
- **The SOPs should change as the office discovers edge cases, or with input from citizens:** The SOPs should not be static, and should evolve to meet the needs of taxpayers, and further the agencies mission. 

# Residential Appeal Standard Operating Procedures

The following guidelines govern the processing of residential appeals at the CCAO.

## Recent sales and selective reappraisal

Many taxpayers say 'I just purchased my home for \$X, and you have assessed me at \$X*y. Why isn't my assesment equal to my recent sale price?' Usually, they say this when y < 1. Setting an estimated property value equal to a recent sale price is called 'selective reappraisal,' and is generally considered bad practice in assessment. 

Selective reappriasal occurres when one property is reassessed differently than others similar to it. When assessments of properties with recent sales are changed to more closely reflect the sale price, they are selectivly re-appraised. Selective reappraisal is unfair because it treats some taxpayers differently than others. Consider two identical homes nearby one another, one recently purchased, the other not. If their assessments differ considerably, they are disuniform. 

It is important to understand that our residential modeling pipeline produces _the most accurate estimates of fair cash value acheivable with the data and technology we have on hand_. All properties, reguardless of recency of sale, receive equally accurate assessments. Selectivly reappraising property arbitrarily treats some properties differently than others.

In addition, selective reappraisal invalidates sales ratio study statistics. The IAAO recommends that all sales ratio studies incorporate tests for selective reappraisal [(pg. 11)](https://www.iaao.org/media/standards/Standard_on_Ratio_Studies.pdf). The CCAO follows that recommendation because meeting IAAO standards on sales ratio studies by selective reappraisal is akin to cheating. 

For these reasons, 

* A recent sale alone is neither sufficient nor necessary evidence to support an argument of over-valuation.

## Characteristics

An appeal based on characteristics argues that the CCAO's physical characteristic data is inaccurate. Inaccurate physical data will likely result in an inaccurate valuation. 

**RAY & MARTY **

## Vacancy

**RAY & MARTY **

## Uniformity

An appeal based on uniformity argues that the appellant's property assessment is inconsistent with assessments of similar properties nearby. The simplest example consiststs of adjacent identical properties; uniformity requires that these properties receive virtually identical assessments. Similar properties that are close to each other should receive similar assessments. As properties become increasingly dissimilar, or more distant geographically, their assessments also become more dissimilar. The degree of dissimilarity required to support a uniformity argument is defined by these SOPs.

### Uniformity comparables

Each property in the County has a set of other properties that form their 'comparable set.' This set is defined as **DAN SNOW**

### Uniformity rule

/> 10/% above the median of the uniformity distribution.

## Valuation

### Sale comparables

**DAN SNOW**

### Overvaluation rule

/> 10/% above the median of the sale comparable distribution.

## Examples

# Managerial overrides



# Updating these SOPs

## Adoption
If these SOPs are approved by the Cook County Assessor, the proposed SOPs will be put to the Valuations Department Residential Tax Analysts to process appeals in 2021. The impact of the SOPs will be analyzed at the end of the year to determine refinement.
